---
title: "ESPM 174A - Homework assignment 1"
author: "Julia Nicholson"
date: "Sep 4th, 2020"
output:
  html_document: default
  pdf_document: default
---

### Instructions
Work on the exercise below, due September 9 (before midnight) on bCourses. 
Please write a **short** answer or interpretation of results (in "your text here"), as well as the R code you used to get there (in "your code here"). 

When you are done you can can knit this script into a PDF using the "Knit" function of the knitr package (see drop down in tool bar). To complete this assignment you will also need the packages we saw in class (tidyr, dplyr, lubridate).

Here is a helpful tutorial by the creator of RMarkdown: 
https://bookdown.org/yihui/rmarkdown/

And here is a link to RStudio cheat sheets for tidyverse packages:
https://rstudio.com/resources/cheatsheets/

And in particular the lubridate package:
https://rawgit.com/rstudio/cheatsheets/master/lubridate.pdf

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE) # knits results only from code chunks
```

### Load packages
```{r code chunk 0}
library(tidyverse) # a universe of tidy data packages
```

### Exercise
### 1) Load the 'airquality' dataset we explored in class, and plot a time series of ozone concentration. What is the span and frequency of the series, and what is the % of NA's?
```{r code chunk 1}
airquality <- datasets::airquality
airquality$Date <- lubridate::as_date(paste(1973, airquality$Month, airquality$Day, sep= ' '))
head(airquality)
o_wout_NAs <- airquality[ !is.na(airquality$Ozone), ]
num_points <- nrow(airquality)
perc_NAs <- round((nrow(airquality) - nrow(o_wout_NAs))/nrow(airquality)*100, digits=0) 
plot_ozone <- ggplot(airquality, aes(Date, Ozone))+
  geom_point()+
  geom_smooth()+
  theme(axis.text.x = element_blank())+
  scale_x_date(date_breaks = "1 week")+
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
#span = 5 months
#frequency = 1 data point/day
plot_ozone
view(perc_NAs)

```
as shown above, span = 5 months, frequency = 30.5 data point/month, and %NAs is 24%.

### 2) Plot an ACF of the ozone time series. What do you see?
```{r code chunk 2}
acf(o_wout_NAs$Ozone, lag.max = 100)
```
There is no significant autocorrelation after about 4 days, so the Ozone is somewhat dependent on the past few days'values for Ozone, but beyond that it is not related.

### 3) Now summarize mean ozone concentration values by week, and plot the resulting time series. How has data completeness (% of non-NA's) improved? 
```{r code chunk 3}
# your code here
airquality$week <- ((airquality$Month - 5)*30 + (airquality$Month %% 2) + airquality$Day) %/%7
#view(airquality)
airqualityS <- airquality %>% 
  group_by(week) %>% # group by category
  summarize(Ozone_mean=mean(Ozone, na.rm = TRUE), 
            Temp_mean=mean(Temp, na.rm = TRUE), 
            Solar_mean=mean(Solar.R, na.rm = TRUE))
summary(airqualityS)
airqualityS
#view(airqualityS)

plot_ozone <- ggplot(airquality, aes(as.factor(week), Ozone))+
  geom_point()+
  geom_boxplot(fill=NA)+
  theme_bw()+
  labs(x="Week")
plot_ozone

plt_O_week <-  ggplot(airqualityS, aes(week, Ozone_mean, group = week))+
  geom_point()+
  geom_boxplot(fill=NA)+
  theme_bw()+
  labs(x="Week")
plt_O_week

```
Data completeness has improved to 0% because we have averaged out by week, and there is at least one data point per week. 

### 4) Is there any trend in the mean weekly ozone series? Would you say it is monotonic? 
```{r code chunk 4}
# your code here
acf(airqualityS$Ozone_mean, lag.max = 30)

cor(airqualityS$week, airqualityS$Ozone_mean, method="pearson")
cor(airqualityS$week, airqualityS$Ozone_mean, method="spearman")
cor(airqualityS$week, airqualityS$Ozone_mean, method="kendall")

cor(airquality$Temp, airquality$Ozone, method = "pearson", use = "complete.obs")
cor(airquality$Temp, airquality$Ozone, method = "spearman", use = "complete.obs")
cor(airquality$Temp, airquality$Ozone, method = "kendall", use = "complete.obs")

cor(airquality$Solar.R, airquality$Ozone, method = "pearson", use = "complete.obs")
cor(airquality$Solar.R, airquality$Ozone, method = "spearman", use = "complete.obs")
cor(airquality$Solar.R, airquality$Ozone, method = "kendall", use = "complete.obs")
```
because the tau values are so small as shown in the cor of week and ozone mean, it doesn't seem like there is a strong trend, although perhaps a slight possitive monotonicity trend.
### 5) Convert the weekly series into a 'matrix' object in the long format (i.e., time over columns).
```{r code chunk 5}
ozone_week <- airqualityS %>% 
  select(week, Ozone_mean)

ozone_week_w <- ozone_week %>% # w is for wide
  pivot_wider(
    id_cols = week,
    names_from = week, # this is the unique index
    values_from = Ozone_mean
  )
head(ozone_week_w)

#long format
ozone_week_l <- ozone_week_w %>% # l is for long
  pivot_longer(
    cols = "0":"21", # wide format columns to pivot back into long format
    names_to = "week", # specify new name for month column
    values_to = "Ozone_mean" # specify new name for ozone column
  )
head(ozone_week_l)
ozone_week_w_mat <- as.matrix(airqualityS$Ozone_mean)
ozone_week_w_mat
```

### Any notes (optional)
If you need to add any additional text, please do so here.