---
title: "Sagehen Groundwater Data Processing"
author: "Jennifer Natali, with help from chatgpt for learning R, dplyr, stats"
date: "2024 October 24"
output:
  pdf_document: default
  html_document: default
---

### Load and Summarize Response Data

1. Upload data (.csv file)
- groundwater level
- TODO: plant greenness from phenopix
- TODO: discharge?

2. Examine the following properties:
- length and frequency of the time series
- completeness of each time series
- descriptive statistics
  - basics: mean, CV, ACF for each variable
  - normality: histogram, qqplot, skewness, kurtosis

```{r code chunk 1, message=FALSE, warning=FALSE}
# Load libraries
library(dplyr)
library(astsa)
library(lubridate)
library(moments) # for skewness and kurtosis testing

# Setup directories and filepaths
home_dir='/Volumes/SANDISK_SSD_G40/GoogleDrive/GitHub/'
repository_dir = paste(home_dir,'sagehen_meadows/', sep='')
groundwater_filepath = paste(repository_dir,'data/field_observations/groundwater/biweekly_manual/groundwater_biweekly_FULL.csv', sep='')
observation_filepath = paste(repository_dir,'data/field_observations/groundwater/biweekly_manual/groundwater_biweekly_observation_spacing.csv', sep='')

# Load groundwater data
groundwater <- read.csv(groundwater_filepath)

# Manage dates and times
groundwater$timestamp <- ymd_hms(groundwater$timestamp)

# Check timestamp formatting
str(groundwater$timestamp)

# Create columns for date and isoweek (starts on Monday)
groundwater <- groundwater %>% mutate(
  date = as.Date(timestamp), 
  year = year(timestamp),
  isoweek = isoweek(date),
  day_of_year = yday(date))

# summarize the full times series
summary(groundwater)

# use z-score? test if data is normally distributed

# shapiro-wilk test; data is likely non-normal if p-value < 0.05
shapiro.test(groundwater$ground_to_water_cm)
# results: p << 0.05, data is non-normal

# Q-Q plots; data is normal if falls on a straight line
qqnorm(groundwater$ground_to_water_cm)
qqline(groundwater$ground_to_water_cm)
# results: mostly normal but some outliers

# Histogram; check for bell-shaped curve
hist(groundwater$ground_to_water_cm)

# Skewness; test if near 0 (symmetric), >0 (positive skew), <0 (neg skew)
skewness(groundwater$ground_to_water_cm, na.rm = TRUE)
# result: 0.83; positive skewed, >0.5 so moderately skewed

# Kurtosis; test for heavy tails (if ~3 normal, if >3 heavy tails + sharp peak, if <3 light tails, flat peak)
kurtosis(groundwater$ground_to_water_cm, na.rm = TRUE)
# result 4.5; heavy tail and sharp peak

# YES, USE Z-SCORE!
# ---TODO: Add ground_to_water_zscore column

# summarize the span of full time series by date
groundwater %>% summarize(
  start_date = min(date, na.rm=TRUE),
  stop_date = max(date, na.rm=TRUE),
  timespan = difftime(stop_date, start_date, units="days"),
  unique_dates_count = n_distinct(date)
)

# summarize the weekly data
groundwater_weekly_summary <- groundwater %>%
  group_by(isoweek) %>%
  summarize(
      n_week = n()  # Number of entries in each week
  )
groundwater_weekly_summary

# summarize spacing of the full time series
unique_observations <- groundwater %>%
  select(-well_id, -ground_to_water_cm, -timestamp) %>%
  distinct(date, .keep_all=TRUE) %>%
  group_by(year) %>%
  arrange(date) %>%
  mutate(day_diff = as.numeric(difftime(lead(date), date, units="days")))

unique_observations %>% summarize(
  max_days = max(day_diff, na.rm=TRUE),
  mean_days = mean(day_diff, na.rm=TRUE),
)

# evaluate how many observation dates are "close" and if they're in the same week
close_threshold = 3 # consider observations "close" if <3 days apart
same_week_count <- unique_observations %>%
  mutate(next_isoweek = lead(isoweek)) %>%
  filter(day_diff < close_threshold) %>%
  summarise(
    close_days_count = n(),
    close_count = sum(isoweek == next_isoweek)
  ) %>%
  mutate(
    percent_close = close_count / close_days_count * 100
  )
same_week_count


# filter measurements, only before 11a
am_time_limit <- 10
groundwater_filter_by_time <- groundwater %>%
  filter(hour(timestamp) < am_time_limit)

# number of rows lost from this filter
nrow(groundwater) - nrow(groundwater_filter_by_time)

# filter for duplicate entries (same well, same week)
groundwater_filter_duplicates <- groundwater_filter_by_time %>%
  group_by(well_id, isoweek) %>%
  distinct(well_id, isoweek, .keep_all = TRUE) %>%
  ungroup()

# compare and validate results
summary(groundwater)
summary(groundwater_filter_by_time)
summary(groundwater_filter_duplicates)

# test completeness of full time series 
# ---TODO: create a ground_to_water_greater_than column for NO WATER readings
#          to capture the depth of the well; this is more info than nothing.

# completeness in terms of entries with NA
# ---NOTE: expect 153 NA entries due to NO WATER readings

groundwater %>%
  summarize(
    na_sum = sum(is.na(ground_to_water_cm)),
    na_percent = 100 * sum(is.na(ground_to_water_cm)) / n())

# basic descriptive statistics across all groundwater readings: mean, CV, ACF
groundwater %>%
  summarize(
    mean_value = mean(ground_to_water_cm, na.rm = TRUE),
    sd_value = sd(ground_to_water_cm, na.rm = TRUE),
    var_value = var(ground_to_water_cm, na.rm = TRUE),
    cv_value = 100 * sd_value / mean_value
  )

groundwater_by_day <- groundwater %>%
  group_by(date) %>%
  summarise(
    mean_value = mean(ground_to_water_cm, na.rm = TRUE),
    sd_value = sd(ground_to_water_cm, na.rm = TRUE),
    var_value = var(ground_to_water_cm, na.rm = TRUE)
  )

# basic descriptive statistics by groupings: mean, CV, ACF for each variable
groundwater_by_well <- groundwater %>%
  group_by(well_id) %>%
  summarise(
    mean_value = mean(ground_to_water_cm, na.rm = TRUE),
    sd_value = sd(ground_to_water_cm, na.rm = TRUE),
    var_value = var(ground_to_water_cm, na.rm = TRUE)
  )
groundwater_by_well

groundwater_by_well_year <- groundwater %>%
  mutate(year = year(timestamp)) %>%
  group_by(well_id, year) %>%
  summarise(
    mean_value = mean(ground_to_water_cm, na.rm = TRUE),
    sd_value = sd(ground_to_water_cm, na.rm = TRUE),
    var_value = var(ground_to_water_cm, na.rm = TRUE),
    .groups = "keep"
  )
groundwater_by_well_year

# Trying to get acf, but not sure if this summarized data means anything
acf(groundwater_by_day$mean_value)


# TODO Next step summary statistics
# ------wells in each meadow group: kiln, east, low
# ------wells from each plant functional type: sedge, willow, mixed herbaceous, pine
# ------wells from each hydrogeomorphic zone: riparian, terrace, fan

# TODO: Next time series to consider
# ------discharge (at one point)
# ------daily precipitation (at one point)
# ------sunlight, aka PAR (at one point)
# ------max, mean daily temperature (at each meadow)

```

### Plots

```{r code chunk 2, warning=FALSE}
# Load libraries
library(ggplot2)
library(tidyr)

# Plot (1): the mean annual groundwater level for all wells for each year
ggplot(groundwater_by_well_year, aes(x = year, y = mean_value, group = well_id)) +
  geom_point() +  # Optional: add points at each data point
  labs(title = "Mean Groundwater Levels Over Time (for each well)",
       x = "Year", y = "Mean Ground to Water (cm)") +
  theme_minimal()

# Plot (2): year-over-year daily time series of mean groundwater level 

# ---Setup dataframe with new columns for year and day_of_year 
groundwater_by_day <- groundwater_by_day %>%
  mutate(
    year = year(date),                # Extract the year from Date
    day_of_year = yday(date)          # Extract the day of year (1-365/366)
  )

# ---Add NA values for days with no measurement (or mean_value)
complete_groundwater_by_day <- groundwater_by_day %>%
  group_by(year) %>%
  complete(day_of_year = 
           min(groundwater_by_day$day_of_year):max(groundwater_by_day$day_of_year), 
           fill = list(mean_value = NA))  # Fill missing days with NA

# ---Plot it!
ggplot(complete_groundwater_by_day, aes(x = day_of_year, y = mean_value, color = factor(year), group = year)) +
  geom_point() +
  geom_smooth() +
  theme_bw() +
  labs(title = "Daily Time Series of Mean Daily Groundwater Level per Year",
       x = "Day of Year",
       y = "Mean Ground to Water (cm)",
       color = "Year") +
  scale_x_continuous(breaks = seq(min(groundwater_by_day$day_of_year), 
                                  max(groundwater_by_day$day_of_year), by = 30)) +  # Customize x-axis (days of year)
  theme_minimal()


```

### Research Questions and Hypotheses

1. How does meadow groundwater vary by season and climate as influenced by elevation, hydrogeomorphic zones, and evapotranspiration rates of plant functional types? 
- Hypothesis: I expect evapotranspiration to drive daily and seasonal groundwater levels with sensitivity to meteorology and day length.

2. What controls plant functional type phenology? 
- Hypothesis: peak productivity and senescence will correlate to groundwater levels as governed by meteorology but moderated by hydrogeomorphic zones and elevation.

3. Does discharge, topography or subsurface character influence groundwater reliability? 
- Hypothesis: I expect that groundwater reliability will correlate to topographic convergence or subsurface boundaries (i.e. differing conductivity).

